---
title: "SIMBA Simulation & Evaluation Example"
author:
-   name: "Jakob Wirbel, Morgan Essex, and Georg Zeller"
    affiliation: "EMBL Heidelberg"
    email: "georg.zeller@embl.de"
date: "Date last modified: 23-05-2022"
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{SIMBA toy example}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{ASCII}
---

# Introduction and setup

This vignette aims to introduce the basic functionality of the SIMBA package and show the structure of the H5 files integral to its use. 

Here and in the manuscript, we use the data and metadata from [Zeevi et al.](https://doi.org/10.1016/j.cell.2015.11.001) as input for different case-control simulation frameworks.

```{r setup}
library("tidyverse") # for data wrangling
library("simbaR")
data("toy_example") # example data
```

In principle, any metagenomic count table can be used as input, so long as it follows the same format.

```{r show.data}
toy.feat[1:3, 1:3]
toy.meta[1:3, 2:4]
```

# 1. Simulate metagenomic data

We begin with setting up some minimal parameters for our simulations. Since we begin with a raw count table, we provide some filtering parameters including prevalence and abundance cutoffs, and a pseudocount value. These will be stored in the .h5 file when the simulation is created. 

`subset_size`, `ab.scale`, and `prev.scale` can be vectors of multiple values, and `prev_scale` is a unique requirement of our signal implantation framework (secondary effect size introducing prevalence shifts between groups in addition to abundance shifts, present in all frameworks).

```{r parameters}
# filtering parameters
ab.cutoff <- 1e-04
prev.cutoff <- 0.05
log.n0 <- 1e-05

# other parameters for the simulations
prop.markers <- 0.1
class.balance <- 0.5
subset_size <- c(50, 100)

# effect sizes
ab.scale <- c(1.0, 2.0)
prev.scale <- 0.2
```

## McMurdie and Holmes framework (multinomial)

We will first create 10 simulations for each combination of effect sizes using the multinomial parametric simulation framework from [McMurdie & Holmes](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1003531).

Expected time on a standard laptop: about 1 minute.

```{r mmh.sim}
# give the simulation a filename
sim.mmh <- 'parametric_example_mmh.h5'

# create simulations
system.time(
  create.data.simulation(feat = toy.feat, 
              meta=toy.meta,
              sim.location = sim.mmh,
              sim.type='cross-section',
              sim.method='McMurdie&Holmes',
              filt.params = list(ab.cutoff=as.numeric(ab.cutoff),
                                 prev.cutoff=prev.cutoff,
                                 log.n0=as.numeric(log.n0)),
              sim.params = list(ab.scale=ab.scale,
                                prop.markers=prop.markers,
                                class.balance=class.balance,
                                repeats=10)))
```

H5 files can be read with the `rhdf5` library in R. Each effect size combination is stored as a 'group' within the file (e.g. ab1_rep1 and ab1_rep2 to represent the first and second repeats of the first abundance shift contained in `ab.scale`).

This results in a total of 10 repeats x 2 effect sizes = 20 simulations.

```{r}
temp <- rhdf5::h5ls(sim.mmh, recursive=FALSE)
temp
```

Finally, we randomly resample each of these groups 10 times each at our desired subset sizes and class balance, and save the indices to the H5 file.

```{r mmh.sim.structure}
# create testing indices
system.time(
  create.test.idx(sim.location=sim.mmh,
                subsets=subset_size,
                repetitions=10))
```

Each group stores the simulated data matrix (features), row and column names, group labels, and the indices of ground truth features (marker_idx). 

When the randomized subsets are generated, a subgroup called test_idx is created under each group, which stores a matrix of the indices for each subset size. In this example, there are 11x50 and 11x100 matrices containing the group labels (header row), and then a row of sample indices for each of the desired 10 repetitions (11 total).

```{r}
temp <- rhdf5::h5ls(sim.mmh)
temp[1:9,]

head(rhdf5::h5read(sim.mmh, 'ab1_rep1/test_idx/subset_50'))
```

## Signal implantation framework (without confounders)

Next, we will do the same simulation and subset generation with our implantation framework. 

Expected time on a standard laptop: about 12 seconds.

```{r signal.implantation.sim}
# give the simulation a filename
sim.implant <- 'implantation_example.h5'

# create simulations
system.time(
  create.data.simulation(feat = toy.feat,
                meta = toy.meta,
                sim.location = sim.implant,
                sim.type ='cross-section',
                sim.method ='resampling',
                filt.params = list(ab.cutoff=as.numeric(ab.cutoff),
                                   prev.cutoff=prev.cutoff,
                                   log.n0=as.numeric(log.n0)),
                sim.params = list(ab.scale=ab.scale,
                                  prev.scale=prev.scale,
                                  prop.markers=prop.markers,
                                  conf=NULL,
                                  class.balance=class.balance,
                                  feature.type='all',
                                  repeats=10)))

# create testing indices
system.time(
  create.test.idx(sim.location=sim.implant,
                subsets=subset_size,
                repetitions=10))
```

The resulting H5 file will be identical, but the group names will reflect the inclusion of a prevalence shift effect size capable in our framework.

Note: abundance shifts of 1.0 and prevalence shifts of 0.0 do not introduce any effect sizes (multiplies counts in one group by 1 or shifts 0% of features between groups, respectively).

```{r}
temp <- rhdf5::h5ls(sim.implant, recursive=FALSE)
setdiff(temp$name, c('original_data', 'simulation_parameters'))
```

# 2. Reality assessment

Once data has been simulated, several attributes are calculated to check the realism of the simulations, for example feature sparsity and variance. 

```{r}
system.time(
  stats <- reality.check(sim.location=sim.implant, 
                         group='ab2_prev1_rep1', 
                         ml=FALSE)
)
map(stats, ~ head(.))
```

PCoAs may also be used to visualize the difference between real and simulated samples.

```{r pcoa.plots}
library("patchwork")

pcoa.mmh <- pcoa.plot(sim.mmh, 'ab2_rep1', 'bray') +
  labs(title = 'Multinomial, McMurdie & Holmes') 
pcoa.implant <- pcoa.plot(sim.implant, 'ab2_prev1_rep1', 'bray') +
  labs(title = 'Signal Implantation')

(pcoa.mmh | pcoa.implant) +
  plot_layout(guides = 'collect') &
  theme(legend.position = 'bottom')
```

# 3. Apply differential abundance tests

To run a differential abundance test, SIMBA uses its `apply.test` function, which works on one H5 group (effect size combination) at a time.

For time purposes, we will only evaluate the n=50 subset sizes. Expected time on a standard laptop: about 30 seconds.

```{r}
temp <- h5ls(sim.implant, recursive=FALSE)
groups <- setdiff(temp$name, c('original_data', 'simulation_parameters'))
groups.sub <- groups[grepl('ab2', groups)]

system.time(
  pvals <- groups.sub %>%
    map(~ apply.test(
      # full simulation file
      sim.location = sim.implant, 
      # which effect size combination
      group = ., 
      # which subset size(s)
      subset = 50,
      # which normalization method (pass = none)
      norm = 'pass', 
      # differential abundance test
      test = 'wilcoxon')) %>%
    flatten()
)
```

`apply.test` returns a matrix of p-values for each subset size, where each row is a bacterial taxon and each column is the significance of the differential abundance of that taxon across each of the 10 repetitions.

```{r}
head(pvals[[1]]$subset_50)
```

# 4. Evaluate test results

To evaluate differential abundance test results, the p-value matrix above can be fed to the `eval.test` SIMBA function, which uses the ground truth stored in marker_idx for each group to tabulate a confusion matrix, precision, recall, AUROC, and a FDR for each element.

```{r}
system.time(
  evals <- groups.sub %>%
    map(~ eval.test(
      # full simulation file
      sim.location = sim.implant,
      # which effect size combination
      group = .,
      # single p-value matrix
      res.mat = pvals[[.]]$subset_50,
      # desired significance threshold
      alpha = 0.05)) %>%
    set_names(groups.sub) %>%
    bind_rows(.id = 'group')
)

evals[1:10,]
```

## Session Info

```{r echo=FALSE, message=FALSE}
file.remove(sim.implant)
file.remove(sim.mmh)
```

```{r}
sessionInfo()
```

